PyTracer: Automatically Profiling Numerical Instabilities in Python

### 2022-05-12 IEEEtran/paper-annotated.tex
Some edits

\newcommand{\cross}[0]{\cellcolor{red!65}\ding{53}}
\newcommand{\valid}[0]{\cellcolor{green!75!black}\ding{51}}
\newcommand{\warn}[0]{\cellcolor{orange!75}?}
\newcommand{\na}[0]{\cellcolor{gray!25}}
\newcommand{\s}[1]{\cellcolor{cyan!25}#1}
\newcommand{\scross}[0]{\ding{53}~}
\newcommand{\svalid}[0]{\ding{51}~}
\newcommand{\swarn}[0]{?~}
\newcommand{\pytracer}[0]{PyTracer\xspace}

@@

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\IEEEtitleabstractindextext{%
    \begin{abstract}
        Numerical stability is a crucial requirement of reliable scientific
        computing. However, despite the pervasiveness of Python in data science,
        analyzing large Python programs remains challenging due to the lack of
        scalable numerical analysis tools available for this language. To fill
        this gap, we developed \pytracer, a profiler to quantify numerical
        instability in Python applications. \pytracer transparently instruments
        Python code to produce numerical traces and visualize them interactively
        in a Plotly dashboard. We designed \pytracer to be agnostic to numerical
        noise model , allowing for tool evaluation through Monte-Carlo
        Arithmetic, random rounding, random data perturbation, or structured
        noise for a particular application. We illustrate \pytracer's
        capabilities by testing the numerical stability of key functions in both
        SciPy and Scikit-learn, two dominant Python libraries for mathematical
        modeling. Through these evaluations, we demonstrate \pytracer as a
        scalable, automatic, and generic framework for numerical profiling in
        Python.
    \end{abstract}

@@

\pytracer estimates the variability of floating-point variables by combining
program traces obtained with different numerical perturbations, which requires
(\add{a}\remove{1}) generating numerically-perturbed executions,
(\add{b}\remove{2}) tracing floating-point computations along program
executions, and (\add{c}\remove{3}) visualizing stability evaluations.
\add{PyTracer uses the ``fuzzy" Python environment instrumented with Monte-Carlo
    arithmetic to address the requirement (a). \mbox{\pytracer} addresses the
    remaining requirements with (b) a dynamic instrumentation of Python functions,
    modules, and classes through meta-programming, (c) an interactive Plotly
    dashboard to visualize stability measures throughout program executions.
    \mbox{\pytracer} is available at
    \mbox{\href{https://github.com/big-data-lab-team/pytracer.git}{github.com/big-data-lab-team/pytracer.git}}
    under MIT License.} \remove{\mbox{\pytracer} addresses these requirements with
    (1) a ``fuzzy" Python interpreter instrumented with Monte-Carlo arithmetic, (2)
    a dynamic instrumentation of Python functions, modules, and classes through
    meta-programming, (3) an interactive Plotly dashboard to visualize stability
    measures throughout program executions. }

@@

\section{Philosophy of \pytracer}

\add{PyTracer follows a three steps approach:}
\begin{enumerate}
    \item \add{Automatic instrumentation of Python programs to save the value of inputs and outputs of function in a file.}
    \item \add{Aggregation of the traces generated to produce a statistics summary.}
    \item \add{Interactive visualization to navigate through the code execution.}
\end{enumerate}

\add{
    The main goal of PyTracer is to help the developper visualizing the numerical
    stability of Python applications. PyTracer saves time for the user by
    instrumenting and aggregating values. In the same way that GDB makes the user
    save time when debugging an application by providing all information in one
    layout, PyTracer allows the user to visualize the overall numerical quality to
    quickly detect instabilities. The PyTracer visualization layout improves result
    analysis in any case since 1) the user does not have to manually instrument its
    code to dump raw data and creates specific post-processing scripts to visualize
    it 2) PyTracer allows the user to export the raw data traced to extend the
    analysis. 
}

\add{
    The main challenge addresses by PyTracer is the transparent instrumentation of Python 
    applications. Indeed, since Python is interpreted language, one has no complete 
    view of the function that will be executed by a given program. 
    Moreover, since we target Python application used in HPC environments, one has
    most of the time no access to source codes or with limited write access.
    Therefore, PyTracer tackles this issue by instrumenting the function dynamically 
    thanks to the Python introspection.
}

\add{
    The based-trace approach used by PyTracer permits to limit the 
    overhead at runtime since the statistics computation is reported to the postprocess analysis.
    However, it limits the aggregation to deterministic executions,
    although control-flow analysis can help to merge divergent execution paths.
}

@@

@@

\pytracer dynamically instruments Python modules without requiring source code
modifications in the application. To do so, it creates a new, instrumented
version of each module that preserves its original attributes so that the
application can transparently use it.
% \gkmod{}{Note: this form of monkey patching is basically global and
% transparent decorating, right?} \Yohan{Yes}
By default, \pytracer instruments all the attributes in a given module except
the special ones of the form \texttt{\_\_*\_\_}, which is common for reserved
Python methods. However, some attributes are not writable. In this case,
\pytracer \remove{preservers} \add{preserves} the original attributes and warns
the user. The user can also restrict the list of traced attributes through an
inclusion/exclusion mechanism. The main advantage of this instrumentation
approach is the lack of required modification in the application source code, in
contrast with \remove{decorator-based approaches} \add{a decorator-based
    approache~\mbox{\cite{hunt2019decorators}}}. In addition, this technique is
scalable and applicable in read-only environments such as Singularity containers
or HPC clusters.

@@

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing to simulate
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision. MCA
can be applied in three modes \add{but we only used the} Random Rounding (RR) \add{in our
    experiments} \remove{, Precision Bounding (PB), and full MCA,} which apply
function $inexact$ to the output. \remove{, the inputs, or both, respectively.
    While the RR mode is equivalent to stochastic rounding, the PB mode can also
    identify catastrophic cancellations.} \add{Random Rounding simulates roundoff errors.}

@@

We enable MCA in Python programs via Verificarlo~\cite{verificarlo}, a
clang-based compiler~\cite{lattner2008llvm} that replaces floating-point
operations by a generic call to a configurable floating-point model. Several
floating-point models are
available~\cite{chatelain2019automatic,chatelain2019outils}, also called
backends. \pytracer leverages ``fuzzy"~\cite{kiar2020comparing}, a collection of
software tools compiled with Verificarlo. In particular, fuzzy provides
MCA-instrumented versions of the Python interpreter as well as the BLAS, LAPACK,
NumPy, SciPy, and scikit-learn libraries, which enables MCA for a wide range of
existing scientific Python programs. Fuzzy is available in Verificarlo's GitHub
organization at
\href{https://github.com/verificarlo/fuzzy}{\url{github.com/verificarlo/fuzzy}}.

@@

Once traces are available, \pytracer sequentially parses them and computes the
mean, standard deviation, and the number of significant digits (as in
Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary
statistics in an HDF5 file, a hierarchical format that facilitates visual
browsing since only required information are loaded when requested by the user.
This operation critically relies on the ordering of call input and output
objects in the traces. Therefore, it assumes that the application is
single-threaded and that the control flow does not depend on a random state.
\add{Therefore, \mbox{\pytracer} cannot merge traces resulting from a different
    control flow path and will stop the parsing when encountering one.}

@@

We evaluated \pytracer on two widely-used Python libraries:
SciPy~\cite{virtanen2020scipy}, the main library for scientific computing in
Python, and  scikit-learn~\cite{pedregosa2011scikit}, a reference
machine-learning framework that implements a wide range of state-of-the-art
algorithms. For each library, we executed the software examples provided in
their respective GitHub repositories. We traced each example five times, having
activated MCA in the Python interpreter \add{as well as} BLAS, LAPACK, NumPy,
SciPy, and scikit-learn. \add{Only instrumented the Python interpreter would
    lead to only perturbe pure Python floating-point operations and thus to miss
    operations in compiled codes (i.e. Cython and shared libraries).} \add{We
    repeated the experiment with two MCA modes, Random Rounding (RR) and Full MCA,
    resulting in 2 experiments for each example}. We used the default virtual
precision in Verificarlo, namely 24 bits for single-precision floating-point
values and 53 bits for double-precision values, which simulates machine error.
We used Python-3.8.5, NumPy-1.19.1, SciPy-1.5.4, and scikit-learn-0.23.2.

\add{Monte Carlo Arithmetic samples the floating-point variability of a given
    program, and hence the more MCA repetitions we have, the more accurate will be
    the sampling distribution. In theory, an infinite number of MCA repetitions
    should give the exact distribution. While in practice low sample size (3$\simeq$30)
    can give a good estimate, the issue is to capture "rare events" that could make the
    computation strongly diverge from the mean. For now, determining the number of
    MCA samples to use is still an open research question. 
    Recent theoretical works}~\cite{sohier2021confidence}
\add{on the significant digits computation allow computing an associated
    confidence interval depending on the sample size and the underlying distribution of the outputs.}

@@

\subsection{Pytracer is applicable to scientific computation libraries of reference}

Table~\ref{tab:pytracer_results_summary} summarizes \pytracer's outcome on SciPy
and scikit-learn examples. Columns \textit{trace} and \textit{parse} report on
the tracing and aggregation steps with two possible outcomes: 1) green for
success, meaning that the execution ended well and no issues were raised by the
application, and 2) red for numerical error in the application, meaning that the
execution raised a Python error before the end. \remove{, and 3)
    orange for errors due to the fuzzy noise injection that are
    not considered numerical errors, meaning that the execu-
    tion raised a Python error but due to a perturbation that
    should not have been introduced (i.e. a floating-point value
    representing an integer).}
Overall, the tracing and parsing of SciPy and scikit-learn examples in Random
Rounding mode was entirely successful for 32/40 examples and showed an average
precision loss of 8 bits. The remaining examples failed due to numerical errors
in the libraries, which we discuss hereafter. 
\remove{The instrumentation with Full MCA is
    more invasive and reduced the number of successful exe-
    cutions to 12/40. Among the 28/40 failed executions, one
    failed due to an actual numerical error (Bayesian Ridge
    Regression example) and the other ones failed due to
    perturbations that should not have been introduced.}

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes
the precision for the main outputs of SciPy and scikit-learn examples. We
measured the precision in significant bits on the final primary example output
by taking the element-wise mean for non-scalar values and, if the example
produces many outputs, the average value among all of them. The numerical
quality across all examples varies from 0 to 52, with an average of 44
significant bits. We note that only one example failed due to a
numerical error (\texttt{face\_recognition}
\remove{for RR and \texttt{Bayesian Ridge Regression} for full MCA}).

@@

\subsection{\pytracer supports the visualization of compound data type}

\add{PyTracer supports native floating point formats as well as compound data type,
    such as complex floating point values, N-dimensional arrays and arbitrary Python
    classes. For the Python classes, PyTracer recursively parses the class attributes to
    extract scalar and N-dimensional arrays floating-points values.}

Figures~\ref{fig:fft1D_inputs} and~\ref{fig:fft1D_outputs} \add{show the
    visualization of the \texttt{fft\_1D} inputs and outputs values. Results are
    complex value encoded on double precision. PyTracer allows visualizing the
    numerical precision of real (blue points) and imaginary (orange points) parts.}
\add{The} top row in Figures~\ref{fig:fft1D_inputs} and~\ref{fig:fft1D_outputs}
show the input mean value while the bottom row shows the standard deviation. The
points with low magnitude in Figure~\ref{fig:fft1D_inputs} correspond to inputs
near 0 when the input of $\sin$ or $\exp$ is close to $k\pi$, $k \in
    \mathbb{Z}$. Indeed, since MCA introduces a slight noise, the result is not
exactly 0. We can see in Figure~\ref{fig:fft1D_inputs} that the maximal noise
introduced by MCA on the inputs is in the order of $10^{-14}$, two orders of
magnitude higher than the $ulp \simeq 10^-16$ for double precision. This slight
noise introduced can be interpreted as white noise on the input signal. We can
see on the bottom row of Figure~\ref{fig:fft1D_outputs} that the frequency noise
is of the same order of of magnitude as the input noise, which is expected. The
two peaks at x=38 and x=562 corresponds to the actual frequencies of the input
signal.

Figure~\ref{fig:spline2d_rr} \add{shows the results of the spline evaluation with
    \texttt{bisplev} from the \texttt{spline\_2D} example. It computes the 2D spline
    interpolation for function $z=(x+y)e^{-6(x^2+y^2)}$, with $(x,y) \in
        [-1,1]\times[-1,1]$. The result is a 2D-array that is natively supported by
    PyTracer. We observe a significant loss of precision in places, which follows
    an interesting pattern that could be further investigated.}

@@

\subsection{\pytracer allows visualizing numerical stability at a fine grained level}

\add{We illustrate on two examples how PyTracer is effective for analyzing
    the numerical stability at a fine grained level: Face recognition and Bayesian Ridge Regression.}
The face recognition example uses eigenfaces (eigenvectors) and Support Vector
Machines (SVMs) to classify faces from the Labeled Faces in the Wild dataset.
This example uses a Principal Component Analysis (PCA) to extract the eigenfaces
and then trains an SVM classifier to predict the name associated with a face.
\add{When executing within fuzzy environment, this example crashed with the following
    Python error raised: \texttt{"SVD did not converge"}.}
This example is the only one to raise an error due to the
non-convergence of the Singular Value Decomposition (SVD) performed in the PCA.
The PCA uses the Randomized SVD~\cite{halko2011finding} method
that first computes a low-rank approximation of the input matrix using a
stochastic algorithm and then uses this approximation as input for the SVD. 
\add{When walking through the code with PyTracer, we see that}
the \texttt{randomized\_svd} scikit-learn function uses the Divide \& Conquer SVD
method \texttt{gesdd}. The LAPACK library has two main methods for computing the SVD:
\texttt{gesdd} and \texttt{gesvd}. The former uses a Divide \& Conquer approach,
while the latter uses a QR decomposition. While both methods are expected to
have the same accuracy~\cite{nakatsukasa2013stable}, the Divide \& Conquer
makes the SVD to not converge with MCA.

Replacing \texttt{gesdd} for \texttt{gesvd} solves the SVD convergence issue.
\pytracer allows to quickly visualize the new precision of the result.
Figure~\ref{fig:randomized_svd_S} shows the precision of the eigenvalues
computed displayed in decreasing order with \texttt{gesvd} \add{in single precision}, 
which varies between 15 and 20
significant bits out of the 24 available in single precision. The classifier
metrics (precision, recall, F1-score, and support) are identical to the IEEE
execution for all MCA samples. \add{We can note that even with using the \texttt{gesvd},
    the precision of the singular values (vectors) is not uniform. 
    Indeed, we can see that the precision of the lowest singular values (vectors) decreases 
    until reaching 0 significant bits.}

The second example, compares a Bayesian Ridge Regression (BRR) to the Ordinary
Least Squares (OLS) estimator on a synthetic dataset and for one-dimensional
regression using polynomial feature expansion. Although the example does not
raise a runtime error, the regression coefficients computed from the fitting are
non-significant. \pytracer traces reveal that the SVD solver used is the root
cause for this error. \add{Indeed, the BayesianRidge model first computes an SVD
    before iteratively updating the coefficients until reaching the convergence
    criterion. However, PyTracer reveals an issue in the singular values computed.}
\add{We can see on Figure}~\ref{fig:BRR_svd} \add{that the last singular values is
    \texttt{NaN} due to numerical error during the SVD although the remaining
    singular values exhibit high numerical precision (between 44 and 50 bits).
    Surprinsingly, this \texttt{NaN} value does not raise issue and is silently
    propagated. Indeed, the stopping criterion compares that the sum of the absolute
    differences between coefficients of two successive iterations but coefficients
    contain \texttt{NaN} since they are computed from singular values. Hence, the
    following condition is never met \texttt{if iter\_ != 0 and
        np.sum(np.abs(coef\_old\_ - coef\_)) < self.tol:} since the comparison with a
    \texttt{NaN} is always false.}

\add{The OLS uses the Least Square to fit the model. PyTracer shows that the
    coefficients fitted are \texttt{NaN} values. By looking at the callgraph, we can
    see that the scipy calls the \texttt{lstsq} method from the
    \texttt{scipy.linalg} package which uses the \texttt{gelsd} LAPACK function.
    Figure}~\ref{fig:OLS_svd} \add{shows the singular values returned by the
    \texttt{lstsq} method shows that one singular value has a \texttt{NaN} value
    that corrupted the calculation of the coefficients. By using the \texttt{gelss},
    PyTracer shows that the \texttt{NaN} value disapeared from the singular values
    (fig.}~\ref{fig:OLS_svd_fixed}\add{) and thus does not contaminated the computation of
    the coefficients that exhibit a precision of 42 bits in average. We can see that
    the full access to all values computed by the application with PyTracer allows
    the user to quickly identify the issue.}

In conclusion, \pytracer and the MCA model were helpful to detect the
instability of the \texttt{gesdd} method mentioned in several GitHub issues,
including the LAPACK repository. Furthermore, it is worth mentioning that a Pull
Request has recently been submitted to let the user select the SVD method and
avoid this kind of issue. Thus, \pytracer can detect instabilities that affect
fundamental scientific computing components such as the LAPACK library. \add{The
    unify view of all traces allows the user to quickly navigate through the
    different numerical precision of variables which saves time for the developer
    compare to the print statement debugging.}
% The Pytracer findings can therefore be extended to many application domains. %
% \tristan{conclusion needs to be discussed. Note to self: generaliser un peu la
% conclusion: pytracer permet de se rendre compte de problemes de stabilite dont
% plusieurs librairies sont dependantes.}

>>>>

\newcommand{\cross}[0]{\cellcolor{red!65}\ding{53}}
\newcommand{\valid}[0]{\cellcolor{green!75!black}\ding{51}}
\newcommand{\warn}[0]{\cellcolor{orange!75}?}
\newcommand{\na}[0]{\cellcolor{gray!25}}
\newcommand{\s}[1]{\cellcolor{cyan!25}#1}
\newcommand{\scross}[0]{\ding{53}~}
\newcommand{\svalid}[0]{\ding{51}~}
\newcommand{\swarn}[0]{?~}
\newcommand{\pytracer}[0]{PyTracer\xspace}
\newcommand{\TG}[1]{\color{red}\textsc{From Tristan}: #1\xspace\color{black}}

@@

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\IEEEtitleabstractindextext{%
    \begin{abstract}
        Numerical stability is a crucial requirement of reliable scientific
        computing. However, despite the pervasiveness of Python in data science,
        analyzing large Python programs remains challenging due to the lack of
        scalable numerical analysis tools available for this language. To fill
        this gap, we developed \pytracer, a profiler to quantify numerical
        instability in Python applications. \pytracer transparently instruments
        Python code to produce numerical traces and visualize them interactively
        in a Plotly dashboard. We designed \pytracer to be agnostic to numerical
        noise model , allowing for tool evaluation through Monte-Carlo
        Arithmetic, random rounding, random data perturbation, or structured
        noise for a particular application. We illustrate \pytracer's
        capabilities by testing the numerical stability of key functions in both
        SciPy and Scikit-learn, two dominant Python libraries for mathematical
        modeling. Through these evaluations, we demonstrate \pytracer as a
        scalable, automated, and generic framework for numerical profiling in
        Python.
    \end{abstract}

@@

\pytracer estimates the variability of floating-point variables by combining
program traces obtained with different numerical perturbations, which requires
(\add{a}\remove{1}) generating numerically-perturbed executions,
(\add{b}\remove{2}) tracing floating-point computations along program
executions, and (\add{c}\remove{3}) visualizing stability evaluations.
\add{PyTracer addresses (a) by relying on the ``fuzzy" Python environment instrumented with Monte-Carlo
    arithmetic, (b) by dynamically instrumenting Python functions,
    modules, and classes through meta-programming, and (c) by visualizing stability measures throughout program executions
    with an interactive Plotly
    dashboard.
    \mbox{\pytracer} is available at
    \mbox{\href{https://github.com/big-data-lab-team/pytracer.git}{github.com/big-data-lab-team/pytracer.git}}
    under MIT License.} \remove{\mbox{\pytracer} addresses these requirements with
    (1) a ``fuzzy" Python interpreter instrumented with Monte-Carlo arithmetic, (2)
    a dynamic instrumentation of Python functions, modules, and classes through
    meta-programming, (3) an interactive Plotly dashboard to visualize stability
    measures throughout program executions. }

@@

@@

\TG{I'm still unsure what to do with this part, it looks out of place.}
\add{PyTracer follows a three steps approach:}
\begin{enumerate}
    \item \add{Automatic instrumentation of Python programs to save traces containing input and output values of all functions.}
    \item \add{Traces aggregation to produce a statistics summary.}
    \item \add{Interactive visualization to navigate through code execution.}
\end{enumerate}

\add{
    The main goal of PyTracer is to help developpers visualize the numerical
    stability of Python applications. PyTracer saves time for the user by
    instrumenting and aggregating values. In the same way that GDB makes the user
    save time when debugging an application by providing all information in one
    layout, PyTracer allows the user to visualize the overall numerical quality to
    quickly detect instabilities. The PyTracer visualization layout improves result
    analysis in any case since 1) the user does not have to manually instrument its
    code to dump raw data and creates specific post-processing scripts to visualize
    it 2) PyTracer allows the user to export the raw data traced to extend the
    analysis. 
}

\add{
    The main challenge addressed by PyTracer is the transparent instrumentation of Python 
    applications. Indeed, since Python is an interpreted language, one has no complete 
    view of the function that will be executed by a given program. 
    Moreover, for Python applications used in HPC environments, one has
    most of the time no access to source codes or with limited write access.
    Therefore, PyTracer tackles this issue by instrumenting the function dynamically 
    thanks to the Python introspection.
}

\add{
    The trace-based approach used by PyTracer permits to limit the 
    overhead at runtime since the statistics computation is reported to the postprocess analysis.
    However, it limits the aggregation to deterministic executions,
    although control-flow analysis can help to merge divergent execution paths.
}

@@

\pytracer dynamically instruments Python modules without requiring source code
modifications in the application. To do so, it creates a new, instrumented
version of each module that preserves its original attributes so that the
application can transparently use it.
% \gkmod{}{Note: this form of monkey patching is basically global and
% transparent decorating, right?} \Yohan{Yes}
By default, \pytracer instruments all the attributes in a given module except
the special ones of the form \texttt{\_\_*\_\_}, which is common for reserved
Python methods. However, some attributes are not writable. In this case,
\pytracer \remove{preservers} \add{preserves} the original attributes and warns
the user. The user can also restrict the list of traced attributes through an
inclusion/exclusion mechanism. The main advantage of this instrumentation
approach is the lack of required modification in the application source code, in
contrast with \remove{decorator-based approaches} \add{a decorator-based
    approach~\mbox{\cite{hunt2019decorators}}}. In addition, this technique is
scalable and applicable in read-only environments such as Singularity containers
or HPC clusters.

@@

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing to simulate
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
\add{In our experiments, we used MCA in Random Rounding (RR) mode, which
simulates roundoff errors by applying function $inexact$ only to the
outputs returned by functions. The Precision Bounding (PB) mode can
identify other types of instabilities, in particular catastrophic
cancellations, however, it is more difficult to apply to large code bases
as it often leads to execution errors that require human intervention.}

@@

We enabled MCA in Python programs via Verificarlo~\cite{verificarlo}, a
clang-based compiler~\cite{lattner2008llvm} that replaces floating-point
operations by a generic call to a configurable floating-point model. Several
floating-point models are
available~\cite{chatelain2019automatic,chatelain2019outils}, also called
backends. \pytracer leverages ``fuzzy"~\cite{kiar2020comparing}, a collection of
software tools compiled with Verificarlo. In particular, fuzzy provides
MCA-instrumented versions of the Python interpreter as well as the BLAS, LAPACK,
NumPy, SciPy, and scikit-learn libraries, which enables MCA for a wide range of
existing scientific Python programs. Fuzzy is available in Verificarlo's GitHub
organization at
\href{https://github.com/verificarlo/fuzzy}{\url{github.com/verificarlo/fuzzy}}.

\add{Monte Carlo Arithmetic samples the floating-point variability of a given
    program, and hence the more MCA repetitions we have, the more accurate the
    distribution will be. In theory, an infinite number of MCA repetitions
    should give the exact distribution. While in practice low sample sizes (3 to 30)
    can give good estimates, the issue is to capture ``rare events" that could make the
    computation strongly diverge from the mean. Determining the number of
    MCA samples to use for a given instrumented code is still an open research question. 
    Recent theoretical works}~\cite{sohier2021confidence}
\add{on the estimation of significant digits allow computing an associated
    confidence interval depending on the sample size and the underlying distribution.}

@@

Once traces are available, \pytracer sequentially parses them and computes the
mean, standard deviation, and the number of significant digits (as in
Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary
statistics in an HDF5 file, a hierarchical format that facilitates visual
browsing since only required information are loaded when requested by the user.
This operation critically relies on the ordering of call input and output
objects in the traces. Therefore, it assumes that the application is
single-threaded and that the control flow does not depend on a random state.
\add{Therefore, \mbox{\pytracer} cannot merge traces resulting from different
    control flow paths and will stop the parsing when encountering such a situation.}

@@

We evaluated \pytracer on two widely-used Python libraries:
SciPy~\cite{virtanen2020scipy}, the main library for scientific computing in
Python, and  scikit-learn~\cite{pedregosa2011scikit}, a reference
machine-learning framework that implements a wide range of state-of-the-art
algorithms. For each library, we executed the software examples provided in
their respective GitHub repositories. We traced each example five times, having
activated MCA in the Python interpreter \add{as well as} BLAS, LAPACK, NumPy,
SciPy, and scikit-learn. \add{Instrumenting only the Python interpreter would
    lead to only perturb pure Python floating-point operations and thus miss
    operations in compiled codes (i.e., Cython and shared libraries).} \add{We
    used MCA in Random Rounding (RR) mode, with 5 repetitions}. We used the default virtual
precision in Verificarlo, namely 24 bits for single-precision floating-point
values and 53 bits for double-precision values, which simulates machine error.
We used Python-3.8.5, NumPy-1.19.1, SciPy-1.5.4, and scikit-learn-0.23.2.

@@

\subsection{\pytracer is applicable to reference scientific computating libraries}

Table~\ref{tab:pytracer_results_summary} summarizes \pytracer's outcome on SciPy
and scikit-learn examples. Columns \textit{trace} and \textit{parse} report on
the tracing and aggregation steps with two possible outcomes: 1) green for
success, meaning that the execution ended well and no issues were raised by the
application, and 2) red for numerical error in the application, meaning that the
execution raised a Python error before the end. \remove{, and 3)
    orange for errors due to the fuzzy noise injection that are
    not considered numerical errors, meaning that the execu-
    tion raised a Python error but due to a perturbation that
    should not have been introduced (i.e. a floating-point value
    representing an integer).}
Overall, the tracing and parsing of SciPy and scikit-learn examples in Random
Rounding mode was entirely successful for 32/40 examples\remove{ and showed an average
precision loss of 8 bits}. The remaining examples failed due to numerical errors
in the libraries, which we discuss hereafter. 
\remove{The instrumentation with Full MCA is
    more invasive and reduced the number of successful exe-
    cutions to 12/40. Among the 28/40 failed executions, one
    failed due to an actual numerical error (Bayesian Ridge
    Regression example) and the other ones failed due to
    perturbations that should not have been introduced.}

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes
the precision for the main outputs of SciPy and scikit-learn examples. We
measured the precision in significant bits on the final primary example output
by taking the element-wise mean for non-scalar values and, if the example
produces many outputs, the average value among them. The numerical
quality across all examples varied from 0 to 52, with an average of 44
significant bits. We note that only one example failed due to a
numerical error (\texttt{face\_recognition}
\remove{for RR and \texttt{Bayesian Ridge Regression} for full MCA}).

@@

\subsection{\pytracer supports the profiling of compound data types}

\add{PyTracer supports native floating point formats as well as compound data types,
    such as complex floating point values, N-dimensional arrays and arbitrary Python
    classes.}
Figures~\ref{fig:fft1D_inputs} and~\ref{fig:fft1D_outputs} \add{show the
    visualization of the \texttt{fft\_1D} input and output values. Results are
    complex values encoded in double precision. PyTracer allows visualizing the
    numerical precision of real (blue points) and imaginary (orange points) parts.}
\add{The} top row in Figures~\ref{fig:fft1D_inputs} and~\ref{fig:fft1D_outputs}
show the input mean value while the bottom row shows the standard deviation. The
points with low magnitude in Figure~\ref{fig:fft1D_inputs} correspond to inputs
near 0 when the input of $\sin$ or $\exp$ is close to $k\pi$, $k \in
    \mathbb{Z}$. Indeed, since MCA introduces a slight noise, the result is not
exactly 0. We can see in Figure~\ref{fig:fft1D_inputs} that the maximal noise
introduced by MCA on the inputs is in the order of $10^{-14}$, two orders of
magnitude higher than the $ulp \simeq 10^-16$ for double precision. This slight
noise introduced can be interpreted as white noise on the input signal. We can
see on the bottom row of Figure~\ref{fig:fft1D_outputs} that the frequency noise
is of the same order of of magnitude as the input noise, which is expected. The
two peaks at x=38 and x=562 corresponds to the actual frequencies of the input
signal.

Figure~\ref{fig:spline2d_rr} \add{shows the results of the spline evaluation with
    \texttt{bisplev} from the \texttt{spline\_2D} example. It computes the 2D spline
    interpolation for function $z=(x+y)e^{-6(x^2+y^2)}$, with $(x,y) \in
        [-1,1]\times[-1,1]$. The result is a 2D-array that is natively supported by
    PyTracer. We observe a significant loss of precision in some regions, which follows
    an interesting pattern that could be further investigated.}

@@

\subsection{\pytracer allows visualizing numerical stability at a fine-grained level}

\add{We illustrate on two examples how PyTracer is effective for analyzing
    numerical stability in complex codes at a fine-grained level: Face recognition and Bayesian Ridge Regression.}
The face recognition example uses eigenfaces (eigenvectors) and Support Vector
Machines (SVMs) to classify faces from the Labeled Faces in the Wild dataset.
This example uses a Principal Component Analysis (PCA) to extract the eigenfaces
and then trains an SVM classifier to predict the label associated with a face.
\add{When executing in PyTracer, this example crashed because the Singular Value Decomposition (SVD) did not converge.}
This example is the only one to raise an error due to the
non-convergence of the SVD performed in the PCA.
The PCA uses the Randomized SVD~\cite{halko2011finding} method
that first computes a low-rank approximation of the input matrix using a
stochastic algorithm and then uses this approximation as input for the SVD. 
\add{When walking through the code with PyTracer, we see that}
the \texttt{randomized\_svd} scikit-learn function uses the Divide \& Conquer SVD
method \texttt{gesdd}. The LAPACK library has two main methods for computing the SVD:
\texttt{gesdd} and \texttt{gesvd}. The former uses a Divide \& Conquer approach,
while the latter uses a QR decomposition. While both methods are expected to
have the same accuracy~\cite{nakatsukasa2013stable}, the Divide \& Conquer
does not converge with MCA.

Replacing \texttt{gesdd} for \texttt{gesvd} solved the SVD convergence issue.
\pytracer allows to quickly visualize the new precision of the result.
Figure~\ref{fig:face_recognition_svd} shows the precision of the eigenvalues
computed displayed in decreasing order with \texttt{gesvd} \add{in single precision}, 
which varies between 15 and 20
significant bits out of the 24 available in single precision. The classifier
metrics (precision, recall, F1-score, and support) are identical to the IEEE
execution for all MCA samples. \add{We can note that even with the \texttt{gesvd} method,
    the precision of the singular values and eigenvectors is not uniform. 
    Indeed, we can see that the precision of the lowest singular values and eigenvectors decreases 
    until reaching 0 significant bits.}

The second example compares a Bayesian Ridge Regression (BRR) to the Ordinary
Least Squares (OLS) estimator on a synthetic dataset and for one-dimensional
regression using polynomial feature expansion. Although the example does not
raise a runtime error, the regression coefficients computed from the fitting are
non-significant. \pytracer traces reveal that the SVD solver used is once again the root
cause for this error. \add{Indeed, the BayesianRidge model first computes an SVD
    before iteratively updating the coefficients until reaching the convergence
    criterion. However, PyTracer shows an issue in the singular values computed.}
\add{We can see in Figure}~\ref{fig:BRR_svd} \add{that the last singular values is
    \texttt{NaN} due to numerical errors in the SVD, although the remaining
    singular values exhibit high numerical precision (between 44 and 50 bits).
    Surprinsingly, this \texttt{NaN} value does not raise an exception and is silently
    propagated. Indeed, the stopping criterion is never verified and the 
    optimization stops after reaching the maximal number of iterations.}

\add{Similarly, for OLS, PyTracer tracers show that the
    regression coefficients fitted are \texttt{NaN} values. By looking at the callgraph, we can
    see that the scipy calls the \texttt{lstsq} method from the
    \texttt{scipy.linalg} package which uses the \texttt{gesdd} LAPACK function.
    Figure}~\ref{fig:OLS_svd} \add{shows that one singular value is a \texttt{NaN} value,
    which corrupted the calculation of the coefficients. When using the \texttt{gesvd} method, 
     the \texttt{NaN} value disapears 
    (Fig.}~\ref{fig:OLS_svd_fixed}\add{), resulting in a precision of 
    42 bits in average in the coefficients. We can see that having
    access to all the values computed by the application with PyTracer allows
    the user to efficiently identify numerical issues.}

In conclusion, \pytracer and the MCA model were helpful to detect the
instability of the \texttt{gesdd} method mentioned in several GitHub issues,
including the LAPACK repository. Furthermore, it is worth mentioning that a Pull
Request has recently been submitted to let the user select the SVD method and
avoid this kind of issue. Thus, \pytracer can detect instabilities that affect
fundamental scientific computing components such as the LAPACK library. \add{The
    unified view of all traces allows the user to quickly navigate through the
     numerical precision of different variables.}
% The Pytracer findings can therefore be extended to many application domains. %
% \tristan{conclusion needs to be discussed. Note to self: generaliser un peu la
% conclusion: pytracer permet de se rendre compte de problemes de stabilite dont
% plusieurs librairies sont dependantes.}
